---
title: "Assignment 01 - Introduction to the `spatstat` package"
# author: "FirstName Lastname"
# date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: textmate
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
```


* * *

Complete the following exercises within the lab period and submit to Canvas before leaving. In addition to the points detailed below, 5 points are assigned to the quality of the annotation and to the 'cleanliness' of the code and resulting pdf document. 

## Exercise 1 -- 5 points

* Import the BC park locations dataset.  -- 0.5 point(s)
* Identify the class of each element within the list.  -- 0.5 point(s)
* Visualise each element within the list using methods appropriate to each object class.  -- 1 point(s)
* Convert the data to a `ppp` object (be sure to include information on regions as marks).  -- 2 point(s)
* Run `plot()` on the resulting `ppp` object. -- 0.5 point(s)
* Briefly describe the dataset. -- 0.5 point(s)

Note: You will need to load the `sp` package and make use of the `as.owin()` function.

```{r}

```


## Exercise 2 -- 2 points

* Refine the figure of the point pattern. You must modify at least 5 graphical parameters. The marks need to be visually distinct.   -- 2 point(s)

```{r}

```


## Exercise 3 -- 5 points

* Create a perspective plot of the elevation image. You must modify at least 10 arguments of the `persp` function and can not use the `viridis()` colour pallet shown in the lab example.  -- 4 point(s) 
* Overlay the park locations.  -- 1 point(s) 

```{r}

```


## Exercise 4 -- 2 points

* Split the elevation image into 5 elevation 'classes'.  -- 0.5 point(s)
* Plot the elevation class image and overlay the park locations.  -- 0.5 point(s)
* Identify which of these elevation classes most parks fall in.  -- 1 point(s)

Note: The `table()` function is useful here.

```{r}

```


## Exercise 5 -- 3 points

* Compute the distance from each park to its nearest neighbour.  -- 0.5 point(s)
* Mark the parks with this information.  -- 0.5 point(s)
* Plot the point pattern with information from these new marks.  -- 1 point(s)
* Identify the most isolated park in BC.  -- 1 point(s)


Note: The park names are stored in the `DATA` object.

```{r}

```

## Exercise 6  -- 3 points

* Calculate the median elevation within British Columbia and at park locations.  -- 1 point(s) 
* Generate a kernel density estimate (KDE) of the distribution of elevation values within the province.  -- 0.5 point(s)
* Generate a KDE of the distribution of elevation values at park locations.  -- 0.5 point(s)
* Create a figure that overlays the two distributions.   -- 0.5 point(s)
* Include a legend. -- 0.5 point(s)
* Do think that the spatial distribution of parks is random with respect to elevation?  -- 0.5 point(s)
* Identify both the highest and lowest elevation parks in BC.  -- 0.5 point(s)

Note: You may need to use the `density()`, `lines()`, and `legend()` functions.

```{r}

```
