---
title: "Lab 01 - Introduction to the `spatstat` package"
#author: "Michael Noonan"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: textmate
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=6)

```

```{r, echo = FALSE}
setwd("~/Dropbox/UBC/Teaching_Material/DATA_589/Labs/01_spatstat")
```


* * *
## Background

A spatial point pattern is a dataset comprised of the locations of 'things' or 'events'. This might be the locations of trees in a forest, road traffic accidents, crimes, incidents of diseases, etc... For these data, the spatial arrangement of the points is the focus of investigation. Depending on your analytical aims, this might be a description of spatial trends in the density of points, relationships with covariates, or so on. The analysis of point patterns can provide key evidence in many fields of research (e.g., ecology, epidemiology, geoscience, astronomy, crime research, cell biology, econometrics).  


In this lab we will:

* Learn how to import point data into the `spatstat` package.
* Learn how to add marks to a point pattern and estimate the observation window.
* Explore ways to visualise point patterns and images.
* Learn how to extract and modify the information contained in `ppp` objects.

* * *
## Importing point data into `spatstat`

At its simplest, a spatial point dataset is comprised of the locations of 'things' or 'events' (i.e., a series of x and y coordinates). In `spatstat`, these data are stored in an object of class `ppp` (i.e., a planar point pattern). Before the analytical tools available within the `spatstat` package can be used, point data need to first be imported and converted into a `ppp` object.

Assuming you have already installed the `spatstat` package, the first step of any analysis is to import the package and dataset(s). Here we will work with point data on synaptic vesicles observed in rat brain tissue. These data were used to support work by Khanmohammadi et al. (2014).



These data are part of the `spatstat` package, and stored as a .txt file in a folder that is generated when the package is installed. We could have loaded these data by calling `data(vesicles)`, but the process described below mimics a more realistic workflow where you would be importing data stored locally on your computer.

```{r, message=FALSE}
#load the spatstat package
library(spatstat)

#Define the file path to the dataset
path <- system.file("rawdata/vesicles/vesicles.txt", package = "spatstat.data")

#Import the vesicles dataset
vesicles <- read.table(path,
                       header = TRUE)

#Visualise the data
plot(y ~ x,
     pch = 16,
     col = "#046C9A",
     data = vesicles)
```

Importing point data into `R` is fairly straightforward and similar to many other workflows, but, unlike with other packages, converting these data to a `ppp` object requires additional information on the sampling window. There are several ways to do this, but we will explore an option based on importing information on the coordinates of a 'bounding box'. This approach assumes you have information on the coordinates defining the edge of a sampling window stored in a data file (e.g., a .txt or .csv file) . The process is fairly straightforward and involves importing the coordinates and converting them to an `owin` object using the `owin()` function. Depending on the complexity of the window, this may involve converting a dataset into a list, as shown in the example below.

```{r, message=FALSE}
# Import the locations of the 
path <- system.file("rawdata/vesicles/vesicleswindow.txt", package = "spatstat.data")

ves_win <- read.table(path,
                      header = TRUE)


# Convert to a list with each element containing information on each "piece"
# This is because there is a hole in the window.
ves_win_stack <- list()
ves_win_stack[[1]] <- ves_win[which(ves_win$id == 1),]
ves_win_stack[[2]] <- ves_win[which(ves_win$id == 2),]


#Convert the list to an owin object
ves_win <- owin(poly = ves_win_stack)

#Visualise the window
plot(ves_win,
     main = "Observation Window")
```

Once the window is defined, converting a dataset into a `ppp` object is relatively straighforward and involves the `ppp()` function.

```{r, message=FALSE}
#Convert to a ppp object
vesicles_ppp <- ppp(x = vesicles$x, # X coordinates
                    y = vesicles$y, # Y coordinates
                    window = ves_win) # Observation window

#Visualise the dataset
plot(vesicles_ppp,
     pch = 16,
     cols = "#046C9A",
     main = "Vesicles point data")
```

### Estimating the window

If you only had point data and no information on the window, the `spatstat` package has methods for estimating the observation window. One option is to use the `ripras()` function, which computes the Ripley-Rasson estimate of the spatial domain from which a particular set of data came.

```{r, message=FALSE}
#Estimate the sampling window
est_win <- ripras(x = vesicles$x, y = vesicles$y)


#Convert to a ppp object
vesicles_ppp_2 <- ppp(x = vesicles$x, # X coordinates
                      y = vesicles$y, # Y coordinates
                      window = est_win) # Observation window

#Visualise the two datasets
par(mfrow = c(1,2))
plot(vesicles_ppp,
     pch = 16,
     cols = "#046C9A",
     main = "True Window")

plot(vesicles_ppp_2,
     pch = 16,
     cols = "#046C9A",
     main = "Estimated Window")
```

While this approach can serve as a reasonable solution for situations when the window is unknown, it risks biasing any downstream estimates, and any resulting inference should be approached with caution.


### Marked point patterns

Sometimes we have points of several types, or a marked point pattern (i.e., auxiliary information). While the original vesicles dataset does not contain any 'marks' we can easily add some randomly generated supporting information for demonstration purposes. 

```{r, message=FALSE}

#Randomly assign a "group"
group <- as.logical(rbinom(n = nrow(vesicles), size = 1, p = 0.5))

#Randomly define a "size"
size <- rgamma(n = nrow(vesicles), shape = 1)

#Assign the supporting information to the point pattern
marks(vesicles_ppp) <- data.frame(Group = group,
                                  Size = size)

#Visualise the result
plot(vesicles_ppp,
     main = "Marked Point Pattern")
```


* * *
## Inspecting and Exploring data

### Plotting methods

#### Plotting a point pattern

As shown above, running plot on `spatstat` objects will generate simple plots of point pattern datasets and features (e.g., marks, windows, etc.). Effective plots of spatial data are critical for communication, and typically requires bespoke modifications of standard, default plotting methods. We will explore some of these options here.

The default `plot` method for `ppp` objects displays the observation window, the points, as well as information on all marks associated with the dataset. For information see `help("plot.ppp")`. The defaults are useful for a quick, 'on-the-fly' visualisation, but are rarely useful for scientific communication. Depending on your needs there is a lot of flexibility in how these figures can be made to look. 

```{r, message=FALSE}
library(viridis)

#Define a colour pallet to use
col_pal <- colourmap(magma(128), range = range(size))

#Refine the figure
plot(vesicles_ppp, # The dataset to visualise
     which.marks = "Size", # Which mark to use
     col = "grey30", #The colour of the window
     cols = col_pal, #The colours of the points
     pch = 16, # The plotting symbol
     main = "Vesicle sites", # The title
     par(bg="grey90", cex.main = 3), # Flexible modification of the graphical parameters
     legend = F) # Turn of the legend depending on needs
```


#### Plotting a window

In some cases we might be interested in the window alone. This can be done by extracting the window from the `ppp` object.

```{r, message=FALSE}
plot(vesicles_ppp$window,
     col = rgb(0,0,0,0.2))
```

The above plot is simple, but it can be modified as needed. See `help("plot.owin")` for details.

#### Plotting an image

Sometimes a point pattern will be accompanied by a continuously varying co-variate. In `spatstat` these covariates are imported as 'images'. Depending on the information contained in these covariates, they can be visualised in two, or three dimensions. By default, the 3 dimensional perspective plots can be challenging to interpret, but they are very flexible and can produce high quality images (see: `help("persp")`). The default plotting methods for 2 dimensional plots of images are redily interpretable, so we will not explore them in detail here. If you want to modify them, see `help("plot.im")`.

To demonstrate how to visualise images, we will use the `bei` dataset. This is a point pattern giving the locations of 3605 trees in a tropical rain forest in Panama. Accompanied by covariate data giving the elevation (altitude) and slope of elevation in the study region. The supporting information is stored in an object called `bei.extra`.

```{r, message=FALSE}
#Load in the data
data("bei")

plot(bei.extra$elev)


persp(bei.extra$elev)

fig <- persp(bei.extra$elev, # source data
             theta = -45, phi = 18, # rotation
             expand = 7, # z-axis expansion
             border = NA, #remove grid borders
             apron = TRUE, #apron around edge
             shade = 0.3, # shading
             box = FALSE, # axes on/off
             main = "", # title
             visible = TRUE, #Supporting calculations
             colmap = viridis(200) ) # colour pallet

perspPoints(bei, Z = bei.extra$elev, M = fig, pch = 16, cex = 0.5)
```

The extra step of setting `visible = TRUE` is required for overlaying the locations of points on top of a perspective plot. This determines which portions of the plot are actually visible, and is then passed on to the `M` argument in `plot.im`.

Sometimes you might be interested in visualising a transect of the values of a supporting covariate across the sampling window. This can be useful way of seeing some of the spatial structure in a covariate. This can be achieved via the `transect.im()` function. Note: by default the transect goes from the bottom left of an image, to the top right, but this can be modified as needed.

```{r, message=FALSE}
plot(transect.im(bei.extra$elev),
     main = "Elevation Transect")
```

Other times you might be interested in dividing a continuously varying image into discrete bins. The `cut.im()` function is a flexible tool for turning a numeric image into a factor-based image. The bin widths are even by default, but can be manually defined to suit your needs.

```{r, message=FALSE}
plot(cut(bei.extra$elev,
         3,
         labels = c("low","medium","high")),
     main = "Elevation classes")
```


The data contained in images can also be passed along to other functions as needed. 

```{r, message=FALSE}
par(mfrow = c(1,2))
hist(bei.extra$elev, main = "")

hist(cut(bei.extra$elev,
         3,
         labels = c("low","medium","high")),
     main = "")
```

### Working with point patterns

#### Extracting information from `ppp` objects

The `spatstat` package has a number of functions for extracting basic information from a point pattern. Some of the more useful ones are described below

```{r, message=FALSE}
#Basic summary information
summary(vesicles_ppp)

#Number of points
npoints(vesicles_ppp)

#marks
head(marks(vesicles_ppp))

#coordinates
head(coords(vesicles_ppp))

#Coordinates and mark information
head(as.data.frame(vesicles_ppp))

```

These functions can also be paired with the assign operator `<-` to modify components of a `ppp` object as needed. For instance, change the `TRUE`/`FALSE` group labels can be done as follows:

```{r, message=FALSE}
#Store the marks
m <- marks(vesicles_ppp)

#Rename as needed
m$Group[which(m$Group == TRUE)] <- "Active"
m$Group[which(m$Group == FALSE)] <- "Inactive"

marks(vesicles_ppp) <- m

head(marks(vesicles_ppp))

plot(vesicles_ppp, which.marks = "Group")
```

This is particularly useful if we calculate values midway through an analyses and would like to append them to our `ppp` object. For example, we can use the `nndist` function to compute the distance from each point to its nearest neighbour. We can then visualise our point pattern based on this additional information.

```{r, message=FALSE}
#Store the marks
m <- marks(vesicles_ppp)

m$Dist <- nndist(vesicles_ppp)

marks(vesicles_ppp) <- m

head(marks(vesicles_ppp))

plot(vesicles_ppp, which.marks = "Dist")
```

Point patterns can be subset using normal `R` data wrangling methods like the `subset` function, or conditional statements. For example, we might be interested in performing our analyses on the points from the "active" group alone.

```{r, message=FALSE}
#Store the marks
m <- marks(vesicles_ppp)

m$Dist <- nndist(vesicles_ppp)

#Choose points from the "active" group
active_ves <- vesicles_ppp[marks(vesicles_ppp)[1] == "Active"]

plot(active_ves, use.marks = FALSE, main = "")
```

#### Working with images

The values contained within images can also be extracted or modified as needed. One of the easiest ways is to use the subset index operator `[]`. For example, if we are interested in identifying the values of an 'image' covariate at the locations where points were recorded we could do this as follows:

```{r, message=FALSE}
#Elevation at tree locations
head(bei.extra$elev[bei])
```

Similarly, a histogram of the elevation at tree locations compared to the elevation across the sampling window is a useful way to visualise whether there is any indication of a non-random spatial distribution of trees.

```{r, message=FALSE}
# histogram of elevations at tree locations overlayed on top of a 
# histogram of elevations within the window
hist(bei.extra$elev,col=rgb(0,0,1,1/4), main = "") #blue
hist(bei.extra$elev[bei], col=rgb(1,0,0,1/2), add = T) # red
```


A full list of all of the functions that can be applied to `spatstat` objects is provided in chapter 4 of Baddeley, Rubak, & Turner (2015).


***
## References

* Baddeley, A., Rubak, E. & Turner, R. (2015). Spatial point patterns: methodology and applications with R. CRC press.

* Khanmohammadi, M., Waagepetersen, R., Nava, N., Nyengaard, J.R. and Sporring, J. (2014) Analysing the distribution of synaptic vesicles using a spatial point process model. 5th ACM Conference on Bioinformatics, Computational Biology and Health Informatics, Newport Beach, CA, USA, September 2014.
