---
title: "Assignment 02 - Exploratory Data Analysis"
# author: "FirstName Lastname"
# date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: textmate
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=6)
```


* * *

Complete the following exercises within the lab period and submit to Canvas before leaving. In addition to the points detailed below, 5 points are assigned to the quality of the annotation and to the 'cleanliness' of the code and resulting pdf document. 

## Exercise 1 -- 2 points

We will again be working with the BC Parks dataset, which contains information on the locations of Provincial Parks in British Columbia. The parks belong to 5 different regions. There is also information on elevation (in m) contained within the dataset.

* Import the BC park locations dataset and convert the data to a `ppp` object (being sure to include information on regions as marks).  -- 0.5 point(s)
* Plot the resulting `ppp` object. The marks need to be visually distinct. -- 0.5 point(s)
* Inspect the spatial distribution of parks. Do you expect the process to be homogeneous? Justify why you came to this expectation. -- 1 point(s)

Note: You will need to load the `maptools` or `sp` packages and make use of the `as.owin()` function.

```{r, message = FALSE}

```


## Exercise 2 -- 2 points

* Under an assumption of homogeneity, what is the intensity of parks/km$^2$ in BC? -- 1 point(s)
* Is the estimated intensity trustworthy? Why/why not? -- 1 point(s)

Hint: see `?rescale`

```{r}

```

## Exercise 3 -- 2 points

* Use a quadrat test to determine whether the assumption of homogeneity is met. -- 1 point(s)
Note: Be sure to set the number of quadrats appropriately, to avoid issues with the quadrat test.
* Visualise both the quadrats and estimated intensity, being sure to include the points in each figure.  -- 1 point(s)
* Is the estimated intensity from exercise 2 trustworthy, and why? -- 1 point(s)

```{r}

```


## Exercise 4 -- 4 points

* Estimate the intensity using kernel estimation with likelihood cross validation bandwidth selection.  -- 1 point(s)
* Perform hotspot analysis to identify locations of elevated intensity.  -- 1 point(s)
* Visualise the output (be sure to include the window).  -- 1 point(s)
* Based on the estimated intensity and hotspot analysis, where would choose to go if you were planning a vacation to tour different provincial parks.  -- 1 point(s)

```{r}

```



## Exercise 5 -- 3 points

* Estimate $\rho$ for the locations of parks as a function of elevation.  -- 1 point(s)
* Plot $\rho$ vs. elevation. Be sure that the x-axis for elevation does not go below 0.  -- 1 point(s)
* Do you think that there is a relationship between elevation and park intensity? Use the results/data to support your statements.  -- 0.5 point(s)
* Would you be more or less likely to find a park at 1500m compared to the average intensity of parks across B.C.? Why?  -- 0.5 point(s)

Note: Estimating rho can be slow ($\sim$ 1-2 min). Be sure to leave enough time for the document to knit.

```{r}

```


## Exercise 6 -- 5 points

* Using Ripley's $K$-function, test for a significant (i.e., $\alpha$ = 0.05) correlation between park locations.  -- 4 point(s)
* Is there any evidence of correlations in park locations? Why? -- 1 point(s)

Notes: Use border corrections (i.e., `correction="border"`) and be sure the estimators assumptions are being met.

```{r}

```


## Exercise 7 -- 3 points

* Using simulation envelopes, estimate both the homogeneous and inhomogeneous pair correlation functions.  -- 1.5 point(s)
* Visualise the results.  -- 0.5 point(s)
* Are the estimates comparable? Which of these would you use to draw conclusions about the clustering of provincial parks? -- 0.5 point(s)
* Are parks in BC clustered? -- 0.5 point(s)

Note: These steps can be slow ($\sim$ 1-2 min). Be sure to leave enough time for the document to knit.

```{r}

```


## Exercise 8 -- 3 points

* Based on these descriptive statistics, what have you learned about the spatial distribution of parks in BC?


